---
title: "STAT 5206: midterm"
author: "de2418"
date: "10/29/2021"
output: html_document
---

<!-- You can skip this and start reading at ### Exam rules -->

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
library(knitr)
library(tidyverse)

# general options
options(
  digits = 3,
  str = strOptions(strict.width = "cut"),
  width = 69,
  tibble.width = 69,
  cli.unicode = FALSE
)

# ggplot options
theme_set(theme_light(base_size = 8))

# knitr options
opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 300,
  out.width = "70%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,
  fig.show = "hold"
)
```

### Exam Rules

- This is an open-book, open-note, open-internet exam.

- You cannot consult another intelligent being for this exam nor should you
  distribute this exam on any platform.

- You need to submit both a RMarkdown file and an HTML of the knitted output:

  - Submissions that do not contain both files will __NOT__ be graded.
  
  - In case your code contains errors, you can add the code chunk option
    `error = TRUE` to ensure that it still knits, as in the example below.
    
    ```{r, error = TRUE}
    stop("I'm an error")
    ```

- Any attempts to print out an entire data frame or list (unless explicitly
  asked for) will be penalized.

- Please knit your file throughout the exam, and upload your solution
  throughout.

- You should not hardcode anything, unless we explicitly tell you it's alright
  to do so.

- The following type of questions will be ignored:

  - Question about the differences between your solution and the backup
    dataset.

  - Question about partial credit or specific grade distributions within the
    problem

  - Questions about installing RStudio or the packages already installed in
    class.

- The backup datasets are not meant to validate your answers and steps have been
  taken to prevent you from doing so. Using the backup datasets will result in a
  small penalty for the question that was meant to generate the dataset. You can
  later remove the use of the backup dataset if you solved the problem, then
  there will be no penalty.

- If you use an external source to solve the final, please cite the URL with a
  comment so we know you grabbed it from online and not from each other.

- Please read the entire exam before asking questions.

### Honor Code (1pt)

We, the undergraduate students of Columbia University, hereby pledge to value
the integrity of our ideas and the ideas of others by honestly presenting our
work, respecting authorship, and striving not simply for answers but for
understanding in the pursuit of our common scholastic goals. In this way, we
seek to build an academic community governed by our collective efforts,
diligence, and Code of Honor.

I affirm that I will not plagiarize, use unauthorized materials, or give or
receive illegitimate help on assignments, papers, or examinations. I will also
uphold equity and honesty in the evaluation of my work and the work of others. I
do so to sustain a community built around this Code of Honor.

Have you read the above honor code?

- To answer this, please create a variable called `i_will_follow_the_honor_code`
  and assign the appropriate `TRUE`/`FALSE` value to it.

- You must also create a character variable named `uni` that contains your UNI.
  This variable has to match your uni at the top.

- __WARNING:__ You will receive 0 points for the exam if you do not get this
  problem correct.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r honor}
i_will_follow_the_honor_code <- TRUE
uni <- "de2418"
```

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

### Question 1: general R programming (20 pts)

#### Question 1.1 (5 pts)

In this question, please answer the following 4 basic questions about the
`mtcars` data frame. To get full mark, each answer should be one line of code!

- Print its first 3 rows:

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 1.1.1}
# Your code goes here!
head(mtcars, 3)
```

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

- What are its dimensions?

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 1.1.2}
# Your code goes here!
dim(mtcars)
```
mtcars is of 32 rows and 11 columns

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

- What are the names of its variables?

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 1.1.3}
# Your code goes here!
names(mtcars)
```
As I see, the variable names are "mpg"  "cyl"  "disp" "hp"   "drat" "wt"   "qsec" "vs"   "am"   "gear" "carb"

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

- What are is the maximum value taken by each of its variables?

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 1.1.4}
# Your code goes here!
mtcars %>% summarise_if(is.numeric, max)
```

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

#### Question 1.2 (6 pts)

- Set the seed to 0.

- Sample

  - 1000 N(0, 1) distributed observations and place them in a 500x2 matrix
    `x0`,

  - 1000 N(5, 1) distributed observations and place them in a 500x2 matrix `x`.

- Bind together the two matrices by row into a 1000x2 matrix `x`.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 1.2}
# Your code goes here!
set.seed(0)
x0 <- matrix(rnorm(1000, 0, 1), nrow = 500, ncol = 2)
x <- matrix(rnorm(1000, 5, 1), nrow = 500, ncol = 2)
x <- rbind(x0, x)
```

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

#### Question 1.3 (3 pts)

Write a function `mse <- function(y1, y2) {}` that returns the mean of the
squared difference between the elements of `y1` and `y2`. Here, you can assume
that `y1` and `y2` can be either vectors or matrices of the same dimension.

Hint: think about what is the expected value of `mse(x0, x1)` ?

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 1.3}
# Your code goes here!
mse <- function(y1, y2){mean((y1 - y2)^2)}
```

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

#### Question 1.4 (6 pts)

Write a function `find_closest <- function(y, centers) {}` where

  - `y` is a numeric vector,

  - `centers` is a list of numeric vectors, each of them having the same length
    as `y`. The function should return an integer corresponding to the member of
    the `centers` list that is the __closest__ to `y` according to the mean
    squared differences.

To check your code, you can use the following:

```{r, eval=FALSE}
find_closest(c(0, 0), list(c(0, 0), c(5, 5))) # returns 1
find_closest(c(2, 2), list(c(0, 0), c(5, 5))) # returns 1
find_closest(c(5, 5), list(c(0, 0), c(5, 5))) # returns 2
find_closest(c(3, 3), list(c(0, 0), c(5, 5))) # returns 2
```

Hint: you can use you `mse` function from before along with a `for` loop, or
better, the appropriate `map_***` variant.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 1.4}
# Your code goes here!
find_closest <- function(y, centers){which.min((map2(y, centers, mse)))}
find_closest(c(0, 0), list(c(0, 0), c(5, 5)))
```


<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

### Question 2: data wrangling and visualization (30 pts)

In this section, you will analyze data from the Olympics games covering the
period 1896-2014. The dataset contains two tables, respectively for the Summer
and Winter games, each containing the following variables.

  - `Year`: the year the competition was held.

  - `Sport`/`Discipline`/`Event`:

      - `Sport`: that which is governed by an International Federation (IF).

      - `Discipline`: a branch of a sport comprising one or more events.

      - `Event`: a competition in a sport or discipline that gives rise to a
        ranking.

  - `Country`: a three letter code indicating the country who won the
    corresponding `Medal` at the corresponding `Event`.

  - `Gender`: can be `"Men"` or `"Women"`.

  - `Medal`: can be `"Gold"`, `"Silved"`, `"Bronze"`.

Given the description above, note that in any given year, an event (e.g., 100M
running) can appear in 6 rows (2 genders times 3 medals).

#### Question 2.1 (3 pts)

- Load the `olympics_summer.csv` as `summer` and `olympics_winter.csv` as
  `winter`:

  - Add a column `Season = "Summer"` or season `Season = "Winter"` to each of
    the two tables.

  - Reorder the columns so that `Year` and `Season` come first, and then
    everything else.

  - Print the first 6 rows of each table.

- Create `both_seasons`, a data frame containing all the data, by binding
  `summer` and `winter` together by row.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 2.1}
# Your code goes here!

#import data
summer <- read.csv("olympics_summer.csv")
winter <- read.csv("olympics_winter.csv")

#tidy data
summer <- summer %>% mutate(Season = "Summer") %>% select(Year, Season, everything())
winter <- winter %>% mutate(Season = "Winter") %>% select(Year, Season, everything())

#print top 6
head(summer, 6)
head(winter, 6)

#combine data frame
both_seasons <- rbind(summer, winter)
```

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

#### Question 2.2 (6 pts)

In this question, you will look at the number of sports/disciplines/events per
seasons and, for each season, the number of sports with the most events.

- Using the `both_seasons` data frame from the previous question, create another
  data frame `events` that contains a single row for every distinct combinations
  of the `Season`, `Sport`, `Discipline` and `Event` variables. Note that, if
  you did not manage to solve the previous question, you can use the backup data
  set `backup_olympics_both_seasons.csv`.

- From `events`, produce two summary tables to separately answer each of the
  following two questions:

    - For each of the two seasons, how many distinct sports, disciplines, and
      events are there?

    - For each of the two seasons, what are the three sports with the most
      events?

- Comment on your results.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 2.2}
# Your code goes here!
# Create event data frame
events <- both_seasons %>% group_by(Season, Sport, Discipline, Event) %>% summarize(n = n())

#Find distinct events number
events %>% group_by(Season) %>% summarize(distinct = n())

#Find top 3 events respectively
events %>% arrange(desc(n)) %>% group_by(Season) %>% slice(1:3)
```
For summer, there are 759 distinct sports, disciplines, and events.
For winter, there are 94 distinct sports, disciplines, and events.
For summer, the three sports with the most events are High Jump, 100M Freestyle Swimming, 10M Platform Diving
For winter, the three sports with the most events are Pairs Figure Skating, Individual Figure Skating, 500M Speed Skating.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

#### Question 2.3 (6 pts)

In this question, you will investigate which ones are the most successful
countries in the post World War II era (i.e., after 1945). Similarly as for the
previous question, you should proceed using the `both_seasons` data frame.

Restricting yourself to data from post World War II competitions:

  - Compute the overall number of bronze, silver, gold, and total medals won by
    each country, irrespectively of the season.

  - Display the figures you obtain for the 10 most successful countries, sorted
    by decreasing number of total medals won. Hint: to make sure that your
    output has only 10 rows, you might have to do some pivoting.

  - Comment on the results.

  - Repeat the exercise using only winter competitions in the post World War II
    era. How do the results change?

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 2.3}
# Your code goes here!

# create a data frame of top 10 countries with medal numbers
medal_num <- both_seasons %>% filter(Year > 1945) %>% group_by(Country) %>% 
            summarize(medal_num = n()) %>% arrange(desc(medal_num)) %>% head(10)
medal_num

# create a data frame of top 10 countries with medal numbers only winter
medal_num_win <- both_seasons %>% filter(Year > 1945 & Season == "Winter") %>% group_by(Country) %>% 
            summarize(medal_num = n()) %>% arrange(desc(medal_num)) %>% head(10)
medal_num_win
```

The results show that USA and URS got the most number of medals for Summer and Winter Olympic after WWII
If only considering winter Olympics, the results change that NOR gets to the top, USA drops to 2nd place and URS drops to 3rd place.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

#### Question 2.4 (7 pts)

In this question, you will study the temporal evolution of the number of
distinct events and medal winning countries. Similarly as for the previous
question, you should proceed using the `both_seasons` data frame.

- Produce a single figure to visualize, for each of the two seasons,

  - the number of distinct events as a function of the year,

  - the number of distinct medal winning countries as a function of the year.

- Hints:

  - You can use any method of your choice to represent many variables on a
    single figure (facetting, shapes, linetypes).

  - Make sure that your figure is complete (e.g., axes and legends properly
    set).

- Comment on the results.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 2.4}
library(ggplot2)

# find the number of distinct events vs the year
distinct_event <- both_seasons %>% group_by(Year, Season, Sport, Discipline, Event) %>% summarize(n = n()) %>% group_by(Year, Season) %>% summarize(distinct_event = n())

# find the number of distinct medal winning countries with the year
distinct_medal_c <- both_seasons %>% group_by(Year, Country, Season) %>% summarize(medal = n()) %>% group_by(Year, Season) %>% summarize(distinct_medal_winning_country = n())

# combine two df
distinct_event_count <- cbind(distinct_event, distinct_medal_c[3])

#plot
ggplot(data = distinct_event_count, mapping = aes(x = Year, y = distinct_event, size = distinct_medal_winning_country, color = Season))+
  geom_point()+
  labs(title = "Number of Distinct Events as a Function of Year",
        xlab = "Year",
        ylab = "Distinct Events",
        size = "Number of Distinct medal Winning Country")
```

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

#### Question 2.5 (8 pts)

In this question, you will study the ratio of the number of distinct events for
women to the number of distinct events for men. Similarly as for the previous
question, you should proceed using the `both_seasons` data frame.

- Using the `both_seasons` data frame, create another data frame `gender` that
  contains one row for every distinct combinations of the `Gender`, `Year`, and
  `Season`, along with an additional column `Count` corresponding to the number
  of distinct `Event` for the combinations. Hint: recall that any combination of
  combinations of `Gender`, `Year`, `Season` and `Event` appears multiple time
  given that there is one row per `Medal`.

- Compute, for each of the two seasons, the ratio of the overall number of
  distinct events for women to the overall number of distinct events for men.
  Hint: start from the `gender` data frame.

- Produce a single figure to visualize, for each of the two seasons, the
  evolution of this ratio as a function of the year. Hint: start again from the
  `gender` data frame.

- Comment on both the overall results and the figure.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

```{r 2.5}
#create data frame gender
gender<- both_seasons %>% group_by(Gender, Year, Season, Event) %>% summarize(n = n()) %>% group_by(Gender, Year, Season) %>% summarize(Count = n())

#create Women count and men count of distinct events by season
Women <- gender %>% group_by(Season) %>% filter(Gender == "Women") %>% summarise(women = sum(Count))
Men <- gender %>% group_by(Season) %>% filter(Gender == "Men") %>% summarise(men = sum(Count))

#Combine and summarize the ratio
gender_modified <- cbind(Women, Men[2])
gender_modified %>% group_by(Season) %>% summarize(ratio = women/men)


##create Women count and men count of distinct events by season and year
Women_year <- gender %>% group_by(Season, Year) %>% filter(Gender == "Women") %>% summarise(women = sum(Count))
Men_year <- gender %>% group_by(Season, Year) %>% filter(Gender == "Men") %>% summarise(men = sum(Count))

#Combine and summarize the ratio
Gender_year <- full_join(Women_year,Men_year)
Gender_year <- Gender_year %>% group_by(Season, Year) %>% summarize(ratio = women/men) %>% drop_na()

#Plot
ggplot(data = Gender_year, mapping = aes(x = Year, y = ratio, color = Season)) +
  geom_point()+
  labs(title = "Ratio of Distinct Events of Women over Men as Function of Year",
       xlab = "Year",
       ylab = "Ratio")
```

Overall, there are more distinct men events than women events. But I notice that in winter, there would generally be more women events than in summer. 
This pattern is supported by the plot. I also notice that there is an increase in number of distinct women events over year.

<!-- !!!!!!!! DO NOT MODIFY THIS LINE !!!!!!!! -->

